---
title: "Drake & Competing Artists: Billboard Hot 100"
author: "Britten"
format:
  html:
    theme: cosmo
---

# Project

For my data science project I want to create a website analyzing Drake’s music career and success over time. I will use data such as Billboard rankings, Spotify streams, album releases, and collaborations to create interactive graphs that show how his music and popularity have changed throughout his career.

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(rvest)
library(DT)
library(plotly)
library(crosstalk)

# Function to scrape Billboard Hot 100 data for any artist
get_artist_chart <- \(artist_slug) {
  url <- str_c("https://www.billboard.com/artist/", artist_slug, "/chart-history/")
  page <- read_html(url)

  formatted_name <- case_match(
    artist_slug,
    "j-cole" ~ "J. Cole",
    .default = str_to_title(str_replace_all(artist_slug, "-", " "))
  )

  tibble(
    lead_artist = formatted_name,
    title = page |> html_elements(".artist-chart-row-title") |> html_text2(),
    artist = page |> html_elements(".artist-chart-row-artist") |> html_text2(),
    debut_date = page |> html_elements(".artist-chart-row-debut-date") |> html_text2() |> mdy(),
    peak_pos = page |> html_elements(".artist-chart-row-peak-pos") |> html_text2() |> as.integer(),
    peak_date = page |> html_elements(".artist-chart-row-peak-date") |> html_text2() |> mdy(),
    weeks_on_chart = page |> html_elements(".artist-chart-row-week-on-chart") |> html_text2() |> as.integer()
  )
}

# Drake and top competing artists in Hip-Hop / Rap on Billboard
comparison_artists <- c(
  "drake", "kendrick-lamar", "j-cole", "future", "travis-scott",
  "kanye-west", "lil-wayne", "eminem", "nicki-minaj", "21-savage"
)

songs <- map(comparison_artists, get_artist_chart) |> 
  list_rbind()
```

## Songs Over the Years

Compare Drake's output with competing artists over time. Use the search bar below to select and compare any artists.

```{r}
#| echo: false
#| message: false
#| warning: false

# Calculate number of new Hot 100 songs per year for each artist
yearly_songs <- songs |> 
  mutate(year = year(debut_date)) |> 
  filter(!is.na(year)) |> 
  count(lead_artist, year, name = "song_count") |> 
  group_by(lead_artist) |> 
  complete(year = full_seq(year, 1), fill = list(song_count = 0)) |> 
  ungroup()

# Set up Crosstalk shared data
shared_yearly <- SharedData$new(yearly_songs, key = ~lead_artist)

# Search bar / selector to pick artists to compare
artist_filter <- filter_select(
  id = "artist_selector",
  label = "Type or select artists to compare:",
  sharedData = shared_yearly,
  group = ~lead_artist
)

# Interactive line plot
line_plot <- plot_ly(
  shared_yearly,
  x = ~year,
  y = ~song_count,
  color = ~lead_artist,
  type = "scatter",
  mode = "lines+markers"
) |> 
  layout(
    xaxis = list(title = "Year", dtick = 2),
    yaxis = list(title = "Hot 100 Entries That Year"),
    hovermode = "x unified"
  )

bscols(
  widths = c(12),
  artist_filter,
  line_plot
)
```

## Interactive Song Catalog

Browse and filter all charted songs across artists.

```{r}
#| echo: false
#| message: false
#| warning: false

# Interactive table with column filters to search and compare artists
datatable(
  songs,
  rownames = FALSE,
  filter = "top",
  colnames = c("Artist", "Song Title", "Credit", "Debut Date", "Peak Position", "Peak Date", "Weeks on Chart"),
  options = list(pageLength = 10, autoWidth = TRUE)
)
```

