---
title: "Drake & The Rap Industry: Billboard Hot 100"
author: "Britten"
format:
  html:
    theme: flatly
    toc: true
    toc-title: "Navigation"
---

# Project Overview

For my data science project, I am analyzing Drake’s music career and commercial longevity compared to his peers in the hip-hop industry. Using data scraped from Billboard's official chart history, this page explores how Drake’s output compares to fellow dominant rap artists from 2009 through 2026.

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(rvest)
library(DT)
library(plotly)

# Function to scrape Billboard Hot 100 data for any artist
get_artist_chart <- \(artist_slug) {
  url <- str_c("https://www.billboard.com/artist/", artist_slug, "/chart-history/")
  page <- read_html(url)

  formatted_name <- if_else(
    artist_slug == "j-cole",
    "J. Cole",
    str_to_title(str_replace_all(artist_slug, "-", " "))
  )

  tibble(
    lead_artist = formatted_name,
    title = page |> html_elements(".artist-chart-row-title") |> html_text2(),
    artist = page |> html_elements(".artist-chart-row-artist") |> html_text2(),
    debut_date = page |> html_elements(".artist-chart-row-debut-date") |> html_text2() |> mdy(),
    peak_pos = page |> html_elements(".artist-chart-row-peak-pos") |> html_text2() |> as.integer(),
    peak_date = page |> html_elements(".artist-chart-row-peak-date") |> html_text2() |> mdy(),
    weeks_on_chart = page |> html_elements(".artist-chart-row-week-on-chart") |> html_text2() |> as.integer()
  )
}

# Drake and top competing artists in Hip-Hop / Rap on Billboard
comparison_artists <- c(
  "drake", "kendrick-lamar", "j-cole", "future", "travis-scott",
  "kanye-west", "lil-wayne", "eminem", "nicki-minaj", "21-savage"
)

songs <- map(comparison_artists, get_artist_chart) |> 
  list_rbind()
```

## Yearly Output Comparison (2009–2026)

The line graph below compares Drake’s annual Hot 100 debuts directly against one competitor at a time. Use the dropdown menu in the upper-right corner of the chart to select a competitor.

```{r}
#| echo: false
#| message: false
#| warning: false

# Calculate accurate Hot 100 debuts per year for each artist (2009-2026)
years_range <- 2009:2026

yearly_songs <- songs |> 
  mutate(year = year(debut_date)) |> 
  filter(!is.na(year), year >= min(years_range), year <= max(years_range)) |> 
  count(lead_artist, year, name = "song_count") |> 
  group_by(lead_artist) |> 
  complete(year = years_range, fill = list(song_count = 0)) |> 
  ungroup()

drake_data <- yearly_songs |> filter(lead_artist == "Drake")

competitors <- c(
  "Kendrick Lamar", "J. Cole", "Future", "Travis Scott",
  "Kanye West", "Lil Wayne", "Eminem", "Nicki Minaj", "21 Savage"
)

# Start plot with Drake as the baseline
line_plot <- plot_ly() |> 
  add_trace(
    data = drake_data,
    x = ~year,
    y = ~song_count,
    name = "Drake",
    type = "scatter",
    mode = "lines+markers",
    line = list(color = "#1E3A8A", width = 3, shape = "spline"),
    marker = list(color = "#1E3A8A", size = 7, line = list(color = "#FFFFFF", width = 1.5)),
    hovertemplate = "<b>Drake</b><br>Year: %{x}<br>Songs Debuted: %{y}<extra></extra>"
  )

# Add traces for each competitor (Kendrick Lamar is visible by default)
for (i in seq_along(competitors)) {
  comp_name <- competitors[i]
  comp_data <- yearly_songs |> filter(lead_artist == comp_name)
  is_visible <- (i == 1)

  line_plot <- line_plot |> 
    add_trace(
      data = comp_data,
      x = ~year,
      y = ~song_count,
      name = comp_name,
      type = "scatter",
      mode = "lines+markers",
      visible = is_visible,
      line = list(color = "#DC2626", width = 2.5, shape = "spline"),
      marker = list(color = "#DC2626", size = 7, line = list(color = "#FFFFFF", width = 1.5)),
      hovertemplate = paste0("<b>", comp_name, "</b><br>Year: %{x}<br>Songs Debuted: %{y}<extra></extra>")
    )
}

# Dropdown menu to switch competitors
buttons <- map(seq_along(competitors), \(i) {
  vis <- c(TRUE, rep(FALSE, length(competitors)))
  vis[i + 1] <- TRUE
  list(
    method = "update",
    args = list(list(visible = vis)),
    label = competitors[i]
  )
})

line_plot |> 
  layout(
    title = list(
      text = "<b>Annual Hot 100 Entries: Drake vs. Peer</b>",
      x = 0,
      font = list(size = 17, color = "#1F2937")
    ),
    plot_bgcolor = "#F9FAFB",
    paper_bgcolor = "#FFFFFF",
    xaxis = list(
      title = "Year",
      dtick = 2,
      gridcolor = "#E5E7EB",
      zerolinecolor = "#D1D5DB"
    ),
    yaxis = list(
      title = "Songs Debuting on Hot 100",
      gridcolor = "#E5E7EB",
      zerolinecolor = "#D1D5DB"
    ),
    hovermode = "x unified",
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.22,
      xanchor = "center",
      yanchor = "top"
    ),
    updatemenus = list(
      list(
        type = "dropdown",
        active = 0,
        x = 1.0,
        y = 1.15,
        xanchor = "right",
        yanchor = "bottom",
        bgcolor = "#FFFFFF",
        bordercolor = "#D1D5DB",
        buttons = buttons
      )
    ),
    margin = list(t = 70, b = 60, l = 50, r = 20)
  )
```

## Interactive Song Catalog

Use the search boxes at the top of each column to filter songs by title, artist, debut date, or peak chart performance.

```{r}
#| echo: false
#| message: false
#| warning: false

datatable(
  songs,
  rownames = FALSE,
  filter = "top",
  class = "hover stripe compact cell-border",
  colnames = c("Artist", "Song Title", "Full Credit", "Debut Date", "Peak Position", "Peak Date", "Weeks on Chart"),
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE
  )
)
```

